Pool Model - A unified Mathematical\\ Modelling Framework for Food Science

### 2025-09-03 paper/pool_model.tex
preliminary finish of introduction

\newacro{ode}[ODE]{Ordinary Differential Equation}
\newacro{glv}[gLV]{generalized Lotka-Volterra}
\newacro{abm}[ABM]{Agent-Based Model}

@@

\noindent
\paragraph{Models for spatial effects; PDEs, \acp{abm}, others}
\begin{itemize}
    \item BSim \cite{Gorochowski2012,Dang2020}
    \item Bacterial Branching patterns \cite{Kawasaki1997,Matsushita1998}
\end{itemize}

\noindent
\paragraph{What is the problem/what is the question? How do we address this?}

>>>>

\newacro{ode}[ODE]{Ordinary Differential Equation}
\newacro{pde}[PDE]{Partial Differential Equation}
\newacro{glv}[gLV]{generalized Lotka-Volterra}
\newacro{abm}[ABM]{Agent-Based Model}

@@

% \paragraph{Models for spatial effects; PDEs, \acp{abm}, others}
% \begin{itemize}
%     \item BSim \cite{Gorochowski2012,Dang2020}
%     \item Bacterial Branching patterns \cite{Kawasaki1997,Matsushita1998}
% \end{itemize}

\noindent
% To incorporate spatial heterogeneity or other effects other modeling approaches have been devised.
% \acp{pde} follow a similar pattern as \acp{ode} but also distribute the described variables through
% space.
When dealing with spatial effects such as heterogeneity, we are required to use more extensive
modeling formulations such as \acp{pde} or \acp{abm}.
The former represent an intermediate step between the population-based \ac{ode} models and the
individually-based \acp{abm} and can be effective, especially when dealing with large crowds of
cells that can be approximated as a continuum.
On the other hand, \acp{abm} excel at describing the individual behaviour of cells which can then
lead to more complex collective phenomena.
Tools such as BSim~\cite{Gorochowski2012,Dang2020} or \texttt{cellular\_raza}~\cite{Pleyer2025} are
capable to describe large amounts of cells, performing various intracellular and extracellular
processes.
The latter has been used to adapt \ac{pde}-based models into an \ac{abm} formulation which describe
bacterial branching patterns~\cite{Kawasaki1997,Matsushita1998}.

% \paragraph{What is the problem/what is the question? How do we address this?}

\noindent
In this work, we introduce a new \ac{ode}-based modeling framework which describes different
bacterial species, their metabolic states and external quantities such as nutrients as individual
pools.
The dynamics of bacterial growth can be understood in terms of transitions between the different
bacterial states i.e. pools.
Our model also allows us to capture interactions between these bacterials states and the external
microenvironment.
Compared against the aforementioned methods, this approach turns out to be less complex with general
applicability, low number of parameters and greater explainability for the underlying biological
reality.
To illustrate the generality of the model, multiple examples are reported.
We analyze the spacial limitations of an interacting two-species model and show that it can be fully
understood on an individual-based level in terms of single-cell processes.
Furthermore, we asses how spatial heterogeneity can impact the shape of the growth curves.
